Retrofitting Effect Handlers onto OCaml

### 2020-10-31 retro-concurrency.tex
Fix static semantics

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
    | (E s) -> (fun k1 -> e1)
    | (F f) -> (fun k2 -> e2)
    | e -> (fun k -> match perform e with
            | v -> continue k v
            | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last pattern in |effc| forwards any unhandled effect to
the outer handler and returns the value and exceptions back.

@@

We present the operational semantics for a core language of effect handlers in
Multicore OCaml. The operational semantics intends to faithfully capture the
semantics of Multicore OCaml effect handlers, and to inform an implementation.
We will highlight the differences whenever we differ from the semantics of the
implementation. We have included an executable version of the dynamic semantics
as an OCaml program in the supplementary material \kc{include hl-semantics.ml}.

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen} for effect handlers, following the abstract machine
semantics of Hillerstrom et al.~\cite{Hillerstrom2020}. The crucial difference
from Hillerstrom et al. is that our stacks are composed of alternating sequence
of OCaml and C stack segments. The program state is captured as configuration
$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
under evaluation, its environment $\env$ and the current stack $\stack$. The
term is either an expression $e$ or a value $v$. The value are integer
constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect begin
performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and, importantly, the OCaml stack $\ostack$ under it. Every OCaml
stack is either an empty stack $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$
with the C stack $\cstack$ under it. Thus the stack is an alternating sequence
of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
frame list $\fl$ is composed of individual frames $r$, which is one of an
argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
position of an application with its environment $\env$, a function frame
$\ffun{v}$ with the value at the function position of an application, and
frames for evaluating the arguments of an arithmetic expression.

>>>>

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
    | (E s) -> (fun k1 -> e1)
    | (F f) -> (fun k2 -> e2)
    | e -> (fun k -> match perform e with
            | v -> continue k v
            | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last pattern in |effc| forwards any unmatched effect to
the outer handler and returns the value and exceptions back.

@@

We present the operational semantics for a core language of effect handlers in
Multicore OCaml. The operational semantics intends to faithfully capture the
semantics of Multicore OCaml effect handlers, and to inform an implementation.
We have included an executable version of the dynamic semantics as an OCaml
program in the supplementary material \kc{include hl-semantics.ml}.

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen} for effect handlers, following the abstract machine
semantics of Hillerstrom et al.~\cite{Hillerstrom2020}. The crucial difference
from Hillerstrom et al. is that our stacks are composed of alternating sequence
of OCaml and C stack segments. The program state is captured as configuration
$\config \coloneqq \configa{\tau}{\env}{\stack}$ with the current term $\tau$
under evaluation, its environment $\env$ and the current stack $\stack$. The
term is either an expression $e$ or a value $v$. The value are integer
constants $n$, continuations $k$, closure $\clos{x}{e}{\env}$, effect being
performed $\effval{l}{k}$ and exception being raised $\exnval{l}$. The
environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
either an empty stack $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
C stack $\cstack$ under it. Thus, the program stack is an alternating sequence
of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
frame list $\fl$ is composed of individual frames $r$, which is one of an
argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
position of an application with its environment $\env$, a function frame
$\ffun{v}$ with the value at the function position of an application, and
frames for evaluating the arguments of an arithmetic expression.
